Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Put myself as last author
In this case, I supervised most of the work so it makes more sense for
me to be last author and Santiago first.

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiago R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina.~e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential and sinusoidal density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 and 0.01 is needed for the exponential and the
sinusoidal density function, respectively.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The minimum values of $D$ that are within the 0.1\% threshold are the $D_\text{linear}$
ones, i.e.~they are the optimal values of the distance-size ratios for $V$ and $g_z$.
These results indicate that the values of $D_\text{linear}$ can be safely extrapolated
to the exponential case.
This conclusion can be understood considering that the two dimensional adaptive
discretization and the density-based algorithms are independent of each other: the first
one splits the tesseroid only in the latitudinal and longitudinal directions, while the
latter only does it in the radial direction.
This leads to the fact that the value of $\delta$ needed to limit the numerical error
due to the density variation is independent of the value of $D$ and vice versa.
That's why the optimal distance-size ratios for the exponential density case is the same
as the ones for the linear density case.

@@

Fig.~\ref{fig:exp-densities} shows the density functions for each value of $b$ and
Fig.~\ref{fig:delta-exponential} shows the resulting relative
differences ($V$ and $g_z$).
In order to improve the visualization of the results, each curve corresponds to the
maximum difference between all shell models for the same value of $b$.
The relative difference for $V$ and $g_z$ falls below the 0.1\% threshold
for each computation grid, shell model and $b$ value if $\delta = 0.1$.
This results imply that a value of $\delta=0.1$ is needed in order to
achieve the 0.1\% accuracy threshold for both the potential and the $g_{z}$, every
shell model, computation grid and value of $b$, in case of exponential density.
These results also show that there is no significant relation between the $b$
value and the numerical error, proving that the density-based discretization algorithm
is capable of limiting the error due to the density variation regardless the variation
of the exponential density.

@@

Fig.~\ref{fig:sine-densities} shows the sinusoidal density functions assigned to every
spherical shell and Fig.~\ref{fig:delta-sine} shows the resulting relative differences
between the analytical and the numerical solutions.
In order to improve the visualization of the results, each curve corresponds to the
maximum difference between all shell models for the same value of $b$.
For this kind of sinusoidal density, a value of $\delta=0.01$ is needed in order to
achieve the 0.1\% accuracy threshold for both the potential and the $g_{z}$, every
shell model, computation grid and value of $b$.
This results also proof that even for a continuous density function with several local
maximum and minimums and inflection points, the density-based discretization algorithm
performs well on computing the gravitational fields that a tesseroid with such density
generates.

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=\linewidth]{figures/neuquen-basin-densities.pdf}
}{
\includegraphics[width=0.5\linewidth]{figures/neuquen-basin-densities.pdf}
}
\caption{
    Linear and exponential densities used to compute the gravitational fields generated
    by a tesseroid model of the Neuquen sedimentary basin.
    The height is defined above the mean Earth radius, and its axis is spanned between
    the deepest and the highest point of the basin.
}
\label{fig:neuquen-basin-densities}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/neuquen-basin.pdf}
\caption{
    Gravitational effects of the Nequ\'en sedimentary basin modelled
    using tesseroids with an exponential density function of depth.
    (a)~Topography of the Neuqu\'en Basin (in km) and its location in South America,
    (b)~thickness of the sedimentary basin \citep[in meters;][]{Heine2007},
    (c)~resulting gravitational potential $V$,
    (d)~resulting vertical component of the gradient ($g_z$),
    calculated at 10km of height over the mean Earth radius.
}
\label{fig:neuquen-basin}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/neuquen-basin-diffs.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model of the
    Neuquen basin with an exponential density contrast and with a homogenenous and
    linear density variation.
    \mbox{(a)-(b)}~Differences on $V$ and $g_z$ between the exponential density model
    and the homogenenous density one,
    \mbox{(c)-(f)}~differences on $V$ and $g_z$ between the exponential density model
    and the linear density one, calculated at 10km of height over the mean Earth radius.
}
\label{fig:neuquen-basin-diffs}
\end{figure*}

@@

\noindent The first term on this equation reproduces the potential generated
by a spherical shell with variable density $\rho(r') = ar'$, while the second
term constitutes the potential generated by a spherical shell with homogeneous
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.

>>>>

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, M.E. Gimenez, and L. Uieda]{
    Santiago R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$,
    Mario E. Gimenez$^{1,2}$, and Leonardo Uieda$^3$ \\
    $^1$CONICET, Argentina.~e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential and sinusoidal density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 and 0.01 is needed for the exponential and the
sinusoidal density function, respectively.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\end{summary}

@@

The minimum values of $D$ that are within the 0.1\% threshold are the $D_\text{linear}$
ones, i.e.~they are the optimal values of the distance-size ratios for $V$ and $g_z$.
These results indicate that the values of $D_\text{linear}$ can be safely extrapolated
to non linear density cases.
This conclusion can be understood considering that the two dimensional adaptive
discretization and the density-based algorithms are independent of each other: the first
one splits the tesseroid only in the latitudinal and longitudinal directions, while the
latter only does it in the radial direction.
This leads to the fact that the value of $\delta$ needed to limit the numerical error
due to the density variation is independent of the value of $D$ and vice versa.
That's why the optimal distance-size ratios for the exponential density case is the same
as the ones for the linear density case.

@@

Fig.~\ref{fig:exp-densities} shows the density functions for each value of $b$ and
Fig.~\ref{fig:delta-exponential} shows the resulting relative
differences ($V$ and $g_z$).
In order to improve the visualization of the results, each curve corresponds to the
maximum difference between all shell models for the same value of $b$.
The relative difference for $V$ and $g_z$ falls below the 0.1\% threshold
for each computation grid, shell model and $b$ value if $\delta$ is lower than 0.1.
This results imply that the optimal value of $\delta$ is 0.1 in case of exponential
density.
They also show that there is no significant relation between the $b$ value and the
numerical error, proving that the density-based discretization algorithm is capable of
limiting the error due to the density variation regardless the variation of the
exponential density.

@@

Fig.~\ref{fig:sine-densities} shows the sinusoidal density functions assigned to every
spherical shell and Fig.~\ref{fig:delta-sine} shows the resulting relative differences
between the analytical and the numerical solutions.
In order to improve the visualization of the results, each curve corresponds to the
maximum difference between all shell models for the same value of $b$.
For this kind of sinusoidal density, a value of $\delta=0.01$ is needed in order to
achieve the 0.1\% accuracy threshold for both the potential and the $g_{z}$, every
shell model, computation grid and value of $b$.
Even in case of a continuous density function with several local maximums and minimums
and inflection points, the GLQ along with the two discretization algorithms reproduce
the gravitational fields with an acceptable degree of accuracy.

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=\linewidth]{figures/neuquen-basin-densities.pdf}
}{
\includegraphics[width=0.5\linewidth]{figures/neuquen-basin-densities.pdf}
}
\caption{
    Linear and exponential densities used to compute the gravitational fields generated
    by a tesseroid model of the Neuqu\'en sedimentary basin.
    The height is defined above the mean Earth radius, and its axis is spanned between
    the deepest and the highest point of the basin.
}
\label{fig:neuquen-basin-densities}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/neuquen-basin.pdf}
\caption{
    Gravitational effects of the Neuqu\'en sedimentary basin modelled
    using tesseroids with an exponential density function of depth.
    (a)~Topography of the Neuqu\'en Basin (in km) and its location in South America,
    (b)~thickness of the sedimentary basin \citep[in meters;][]{Heine2007},
    (c)~resulting gravitational potential $V$,
    (d)~resulting vertical component of the gradient ($g_z$),
    calculated at 10km of height over the mean Earth radius.
}
\label{fig:neuquen-basin}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/neuquen-basin-diffs.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model of the
    Neuqu\'en basin with an exponential density contrast and with a homogeneous and
    linear density variation.
    \mbox{(a)-(b)}~Differences on $V$ and $g_z$ between the exponential density model
    and the homogeneous density one,
    \mbox{(c)-(f)}~differences on $V$ and $g_z$ between the exponential density model
    and the linear density one, calculated at 10km of height over the mean Earth radius.
}
\label{fig:neuquen-basin-diffs}
\end{figure*}

@@

\noindent The first term on this equation reproduces the potential generated
by a spherical shell with variable density $\rho(r') = ar'$, while the second
term constitutes the potential generated by a spherical shell with homogeneous
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
Eq~\ref{eq:shell-pot-linear} is in agreement with the one obtained by \citet{Lin2018}.
